---
title: "molic: Multivariate OutLIerdetection In Contingency tables"
output:
  github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warnings = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

[![R build status](https://github.com/mlindsk/molic/workflows/R-CMD-check/badge.svg)](https://github.com/mlindsk/molic/actions)
[![status](https://joss.theoj.org/papers/9fa65ced7bf3db01343d68b4488196d8/status.svg)](https://joss.theoj.org/papers/9fa65ced7bf3db01343d68b4488196d8)
[![DOI](https://zenodo.org/badge/177729633.svg)](https://zenodo.org/badge/latestdoi/177729633)

## About molic

An **R** package to perform outlier detection in contingency tables (i.e. categorical data) using decomposable graphical models (DGMs); models for which the underlying association between all variables can be depicted by an undirected graph. **molic** also offers algorithms for fitting undirected decomposable graphs. Compute-intensive procedures are implemented using [Rcpp](http://www.rcpp.org/)/C++ for better run-time performance.

## Installation

You can install the current stable release of the package by using the `devtools` package: 

```{r, eval = FALSE}
devtools::install_github("mlindsk/molic", build_vignettes = FALSE)
```

## Getting Started

See the following:

 - [The Outlier Model](https://mlindsk.github.io/molic/articles/outlier_intro.html): The "behind the scenes" of the model.
 - [Detecting Skin Diseases](https://mlindsk.github.io/molic/articles/dermatitis.html): An example of using the outlier model to detect skin diseases. 
 - [Outlier Detection in Genetic Data](https://mlindsk.github.io/molic/articles/genetic_example.html): An example of how to conduct an outlier analysis in genetic data.

The following examples are from the documentation of `fit_outlier`.

```{r}
library(dplyr)
library(molic)
library(ess)   # For the fit_graph function
```

Psoriasis patients

```{r}
d <- derma %>%
  filter(ES == "psoriasis") %>%
  select(-ES) %>%
  as_tibble()
```

Fitting the interaction graph

```{r}
g <- fit_graph(d, trace = FALSE) # see package ess for details
plot(g, vertex.size = 15) 
```

This plot shows how the variables are 'associated' in the psoriasis class; see [ess](https://github.com/mlindsk/ess) for more information about `fit_graph`. The outlier model exploits this knowledge instead of assuming independence between all variables (which would clearly be a wrong assumption looking at the graph). The graph may look very different for other classes than psoriasis.

## Example 1 - Testing which observations within the psoriasis class are outliers

```{r}
m1 <- fit_outlier(d, g)
print(m1)
outs  <- outliers(m1, d)
douts <- d[which(outs), ]
douts
```

Notice that m1 is of class 'outlier'. This means, that the procedure has tested which observations _within_ the data are outliers. This method is most often just referred to as outlier detection. The following plot is the distribution of the test statistic. Think of a simple t-test, where the distribution of the test statistic is a t-distribution. In order to conclude on the hypothesis, one finds the critical value and verify if the test statistic is greater or less than this.

```{r}
plot(m1) 
```

Retrieving the test statistic for the individual observations:

```{r}
x1 <- douts[1, ] %>% unlist()
x2 <- d[1, ] %>% unlist()
dev1 <- deviance(m1, x1) # falls within the critical region in the plot (the red area)
dev2 <- deviance(m1, x2) # falls within the acceptable region in the plot
dev1
dev2
```

Retrieving the p-values:

```{r}
pval(m1, dev1)
pval(m1, dev2)
```

## Example 2 - Testing if a new observation is an outlier
 
An observation from class "chronic dermatitis": 

```{r}
z <- derma %>%
  filter(ES == "chronic dermatitis") %>%
  select(-ES) %>%
  slice(1) %>%
  unlist()
```

Test if z is an outlier in class "psoriasis":

```{r}
m2 <- fit_outlier(d, g, z)
print(m2)
plot(m2) # The vertical dotted line indicates the (deviance) test statistic of z
```

Notice that m2 is of class 'novelty'. The term _novelty detection_ is sometimes used in the litterature when the goal is to verify if a new unseen observation is an outlier in a homogeneous dataset. Retrieving the test statistic and p-value for `z`

```{r}
dz <- deviance(m2, z)
pval(m2, dz)
```

## How To Cite

If you want to cite the **outlier method** please use

```latex
@article{lindskououtlier,
  title={Outlier Detection in Contingency Tables Using Decomposable Graphical Models},
  author={Lindskou, Mads and Svante Eriksen, Poul and Tvedebrink, Torben},
  journal={Scandinavian Journal of Statistics},
  publisher={Wiley Online Library},
  doi={10.1111/sjos.12407},
  year={2019}
}
```

If you want to cite the **molic** package please use

```latex
@software{lindskoumolic,
  author       = {Mads Lindskou},
  title        = {{molic: An R package for multivariate outlier 
                   detection in contingency tables}},
  month        = oct,
  year         = 2019,
  publisher    = {Journal of Open Source Software},
  doi          = {10.21105/joss.01665},
  url          = {https://doi.org/10.21105/joss.01665}
}
```

<!-- Also, see the [jti](https://github.com/mlindsk/jti) package which is used for making inference in Bayesian networks or DGMs; in the latter case, one can exploit the **ess** package. -->
