---
title: "molic: Multivariate OutLIerdetection In Contingency Tables"
output:
  github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

[![Build Status](https://travis-ci.com/mlindsk/molic.svg?token=AuXvB5mAnHuxQxKszxph&branch=master)](https://travis-ci.com/mlindsk/molic)

## About molic

An **R** package to perform outlier detection in contingency tables using decomposable graphical models (DGMs); models for which the underlying association between all variables can be depicted by an undirected graph. **molic** also offers algorithms for fitting undirected decomposable graphs. Compute-intensive procedures are implementet using [Rcpp](http://www.rcpp.org/)/C++ for better run-timer performance.

## Getting Started
If you want to learn the "behind the scenes" of the model it is recommended to go through the [vignette tutorial](https://mlindsk.github.io/molic/articles/) and look at the [documentation](https://mlindsk.github.io/molic/reference/index.html) as you read along. 


## Installation
You can install the development version of the package by using the `devtools` package:

```{r, eval = FALSE}
devtools::install_github("mlindsk/molic", build_vignettes = TRUE)
```

## Main Functions
A list of some core functions in the **molic** package is listed below

```{r molic_funcs, echo = FALSE}
func_names <- c(
  "`outlier_model`",
  "`p_val`",
  "`efs`",
  "`cl_tree`")
func_descr <- c(
  "Fits an outlier detection model",
  "Calculates the p-value of an observation being an outlier",
  "Fits a decomposable graph using forward selection.",
  "Fits a tree (decomposable graph) using the Chow-Liu method."
)
func_tab <- data.frame(Function = func_names, Description = func_descr)
pander::panderOptions('table.alignment.default', 'left')
pander::pander(func_tab)
```

## Example
TBA

## How To Cite
TBA
