---
title: "Outlier Detection in Genetic Data"
author: "Mads Lindskou"
output: 
  rmarkdown::html_vignette:
    highligh: zenburn
    toc: true
    fig_height: 4
    fig_width: 4
    fig_align: "center"
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Outlier Detection in Genetic Data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, include = TRUE)
```


# The 1000 Genomes Project Data

The outlier detection method [@lindskououtlier] arose from a problem in the forensic science community where it is of great interest to make statements about the geographical origin of a DNA sample. This is in general a very complicated matter. More over, when DNA markers are in linkage disequilibrium things get even more complicated. The 1000 Genomes Project set out to provide a comprehensive description of common human genetic variation by applying whole-genome sequencing to a diverse set of individuals from multiple populations [@10002015global]. In the **molic** package we include the final data from the project which includes $2504$ DNA profiles coming from five different continental regions; Europe (EUR), Africa (AFR), America (AMR), East Asia (EAS), South Asia (SAS). Each DNA profile is the compound of two rows in the data `tgp_dat`, one for each chromosome copy. This makes the outlier procedure slightly more complicated since we must fit a model to each copy and aggregate the information (but we shall see in a moment that it is not hard to do using the **molic** package).

The data includes $276$ SNP markers grouped in $97$ microhaplotypes; in other words, the SNPs within a microhaplotype are so close that they cannot be assumed to be in linkage disequilibrium and we must take into account their mutal dependencies. We do this with the use of decomposable graphical models. See the `outlier_intro` for more details on the model. The microhaplotype grouping is provided is the list `tgp_haps`.

# An Outlier Test

We show how to test that an unknown DNA profile `z` can be assumed to originate from the European database. We shall pick a random DNA profile from the American database and assume we do not know the origin. 


```{r e}
library(molic)
library(dplyr)

# The components - here microhaplotypes
haps <- tgp_haps

# All the Europeans
eur  <- tgp_dat %>%
  as_tibble() %>%
  filter(pop_meta == "EUR")

# Extracting the two databases for each copy of the chromosomes
eur_a <- eur %>%
  filter(grepl("a$", sample_name)) %>%
  select(-c(1:2))

eur_b <- eur %>%
  filter(grepl("b$", sample_name)) %>%
  select(-c(1:2))

mat_eur_a <- eur_a %>% as.matrix()
mat_eur_b <- eur_b %>% as.matrix()

# Fitting the interaction graphs on the EUR data 
ga <- fit_components(eur_a, comp = haps, trace = FALSE)
gb <- fit_components(eur_b, comp = haps, trace = FALSE)
# print(ga)
# plot(ga, vertex.size = 1)

# Testing if an American is an outlier in Europe

amr <- tgp_dat %>%
  as_tibble() %>%
  filter(pop_meta == "AMR")

z1  <- amr %>%
  filter(grepl("a$", sample_name)) %>% 
  select(unname(unlist(haps))) %>%
  dplyr::slice(1) %>%
  unlist()

z2  <- amr %>%
  filter(grepl("b$", sample_name)) %>% 
  select(unname(unlist(haps))) %>%
  dplyr::slice(1) %>%
  unlist()


# The following takes a couple of minutes
m1 <- fit_outlier(mat_eur_a, z1, ga) # consider using more cores (ncores argument)
m2 <- fit_outlier(mat_eur_b, z2, gb) # consider using more cores (ncores argument)
m  <- fit_mixed_outlier(m1,m2)
```

```{r}
print(m)
plot(m)
```
Thus, `z` does not deviate significantly from the Europe database. The red area depicts the $0.05$ significance level and the dotted line is the observed deviance of `z` (`r m$dev`). This is not a surprise, since Americans and Europeans have many similarities.

# References
