---
title: "The Outlier Model"
author: "Mads Lindskou"
output: 
  rmarkdown::html_vignette:
    highligh: zenburn
    toc: true
    fig_height: 4
    fig_width: 4
    fig_align: "center"
    includes:
      before_body: style.html
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{The Outlier Model}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Background Theory on The Outlier Detection Model

### Contingency Tables
The outlier detection model can only be used on discrete data; i.e data for which all variables can only take on a finite set of values (also called _levels_). We also say, that such variables are _discrete variables_ or _categorical variables_. 

<!-- Examples of such variables are  -->
<!-- ```{r echo = FALSE} -->
<!-- Variables <- c("gender", -->
<!--   "seasons", -->
<!--   "DNA markers", -->
<!--   "grades", -->
<!--   "race", -->
<!--   "treatment" -->
<!-- ) -->
<!-- Levels  <- c("male/female", -->
<!--   "spring/summer/autumn/winther", -->
<!--   "A/T/C/G", -->
<!--   "pass/fail", -->
<!--   "white/hispanic/black", -->
<!--   "active/placebo" -->
<!-- ) -->
<!-- discrete_examples <- data.frame(Variables, Levels) -->
<!-- pander::pander(discrete_examples) -->
<!-- ``` -->

Consider for example the built in discrete data set in `R` on `4526` applicants to graduate school at Berkeley for the six largest departments in 1973 classified by admission and sex. Below, a small sample of the data is seen.

```{r admission, echo = FALSE, message = FALSE, warning = FALSE}
## https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/UCBAdmissions.html
library(dplyr)
adm_ <- as.data.frame(UCBAdmissions) 
adm <-data.frame(Admit = c(), Gender = c(), Dept = c())
for( i in 1:nrow(adm_) ) {
  freq_i <- adm_[i, "Freq"]
  for ( j in 1:freq_i ) {
    adm <- rbind(adm, adm_[i, c("Admit", "Gender", "Dept")])
  }
}
adm <- adm[sample(nrow(adm)), ] %>%
  as_tibble() %>%
  mutate_all(.funs = as.character)
pander::pander(head(adm))
```

Another and more appropriate way of showing the entire data, is via a _contingency table_ which summarizes the counts of all combinations of `Admit`, `Gender` and `Dept`

```{r cont_adm, echo = FALSE}
pander::pander(ftable(adm))
```

This is much more informative and we can e.g. say immediately that `207` `Male` applicants were `Rejected` at department `B`. A particular entry in the table is called a `cell`. So `(Male, Rejected, B)` is a cell with _cellcount_ `207`. The _dimension_ of the table is the number of variables (here `3`). For high-dimensional tables things get more complicated and it may not be feasible to show the table (think of hundreds of DNA markers) and there may be many cells with zero cellcount.

The outlier detection method (described shortly) uses such contingency table data. However, it should be mentioned one more time, that a contingency table is in one-to-one correspondence with a "regular" discrete data set; contingency tables are nothing more than a convenient way of describing discrete data sets.

### Outliers
An outlier can be regarded as an observation which deviates so much from the other observations in a database as to arouse suspicions that it was generated by a different mechanism. The outlier detection method directly adapts this definition by specifying a hypothesis of an outlier being distributed differently than all other observations in a given database. An outlier is also a very case-specific unit, since it may be interpreted as natural extreme noise in some applications and in other applications it might be the most interesting observation. 

Outlier detection is usually divided into \textit{univariate methods} where the task is to detect observations that are extreme for a single variable and \textit{multivariate methods} in which the task is to detect unusual combinations between several variables. 

For high-dimensional data, it is a tedious task to determine anomalies such as outliers. The method described here is a multivariate method for outlier detection in high-dimensional contingency tables, i.e. data with discrete variables only. The method relies on the class of decomposable graphical models to model the relationship among the variables of interest, which can be depicted by an undirected graph called the interaction graph.

### Decomposable Graphical Models
In order to model the interaction structure between variables, we need to construct a _decomposable graphical model_. A graphical model is a statistical model for which an undirected graph represents the interaction between the vertices in the model. An undirected graph is a pair $G = (V,E)$ where $V$ is a set of vertices and $E$ is a set of edges connecting elements in $V$. An edge connecting two vertices indicates that these two are dependend on each other; this is also called a two-way interaction. A threeway interaction occurs when three vertices are all mutually connected (e.g.\ can be visualized as a triangle). An undirected graph is _decomposable_ if there are no cycles of length greater than four without a chord (an edge between two non-adjacent vertices in the cycle). The _subgraph_ $G_{A} = (A, E_{A})$ consist of vertices $A\subseteq V$ from $G$ and the corresponding edges $E_A$ between them. A graph is _complete_ if there is an edge between all pairs of vertices and a complete subgraph is called a (maximal) _clique_ if it is not contained in any other complete subgraph. A subset of vertices is complete if it induces a complete subgraph. Two sets $A,B \subseteq V$ are separated by a third set $C \subseteq V$ if all paths between vertices in $A$ and $B$ go through $C$. If $C$ is the smallest set such that $A$ and $B$ are separated, we say that $C$ is a (minimal) _separator_ for $A$ and $B$.

\

Consider the undirected graph $G$ in the figure below. 

\

```{r undirected_graph, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center", fig.cap = "\\label{fig:DG} $G$: An undirected decomposable graph."}
suppressMessages(library(igraph))
el <- matrix( c("a", "b",
  "b", "c",
  "b", "d",
  "d", "e",
  "d", "c"),
  nc = 2,
  byrow = TRUE
)
G <- graph_from_edgelist(el, directed = FALSE)
set.seed(7)
plot(G, layout = layout_nicely(G), 
     edge.arrow.size=1, 
     vertex.label.cex=1.5, 
     vertex.label.family="Helvetica",
     vertex.label.font=1,
     vertex.shape="circle", 
     vertex.size=40, 
     vertex.label.color="blue", 
     edge.width=5)
```

\

The set of vertices is $V = \{a,b,c,d,e\}$ and the set of edges is \(E = \{ab, bc, bd, cd,de\}\). The cliques are $C_1 = \{a, b\}$, \(C_2 = \{b, c, d\}\) and $C_3 = \{d, e\}$. The minimal separators are $S_{2} = \{b\}$ and $S_3 = \{d\}$ where $S_{2}$ separates $C_{1}$ and $C_{2}$ and $S_{3}$ separates $C_{2}$ and $C_{3}$. Since $G$ has no cycles of length greater than three, the graph is decomposable. Notice, that $a$ is only connected to $\{c,d,e\}$ through $\{b\}$; we interpret this as "$a$ is independent of $\{c,d,e\}$ when we know the value of $b$". Such statements can be used to gain insight into complex structures.

Finally, we can associate a probability measure with an _interaction graph_; an undirected graph with each vertex being a random variable. For decomposable graphs, the probability density function can be written in terms of the cliques and separators. Collectively, models for which the interaction graph is decomposable are called decomposable graphical models (DGM). For more details on graphical models, see for example [@whittaker2009graphical; @lauritzen1996graphical].

This setup allow us to estimate probabilities of having a specific combination of associated variables. Think of the vertex set $V = \{a,b,c,d,e\}$ as a set of five variables where each variable can be either zero or one. Then given the decomposable interaction graph $G$ we can, as an example, estimate the probability that $a = 1, b = 0, c = 1, d = 1$ and $e = 0$. In this contrived example there are $2^{5} = 32$ possible combinations.

Tree graphs are per definition decomposable graphs. Notice that the graph $G$ is not a tree because of the presence of the cycle $\{b,c,d\}$. However, the graph is indeed decomposable.


# The Outlier Detection Model
The method is described using an example. For more details see the original paper [TBA]. We use DNA data from the 1000 Genomes Project [@10002015global] consisting of $2504$ DNA profiles from $5$ different continential regions and $302$ binary DNA markers. The data is shipped with the package and can be used by calling `tgp_dat`. All DNA markers are further grouped into so-called microhaplotypes. DNA markers within a microhaplotype are known to be associated which is perfect to showcase the model. 

### Toy Example Facilitating Theory

For simplicity, we focus on a single microhaplotype containing $3$ DNA markers which we rename to $X, Y$ and $Z$ to have shorter names.

\

```{r message=FALSE, warning=FALSE}
library(molic)

vars <- structure(tgp_haps[[29]], names = c("X", "Y", "Z"))

df_EUR <- tgp_dat %>%
  filter(pop_meta == "EUR") %>%
  select(!!vars)

df_AFR <- tgp_dat %>%
  filter(pop_meta == "AFR") %>%
  select(!!vars)
```

`molic` has a nice function `sptable` for creating sparse contingency tables. The input is a `matrix`. It is sparse in the sense, that it do not show the combinations with zero counts. The European table is

```{r EUR_tab, include = FALSE}
EUR_tab <- sptable(as.matrix(df_EUR))
```

```{r}
sptable(as.matrix(df_EUR))
```

```{r AFR_tab, include = FALSE}
AFR_tab <- sptable(as.matrix(df_AFR))
```

and the African table is

```{r}
sptable(as.matrix(df_AFR))
```

From the EUR table we see that only `r unname(EUR_tab["GGA"])` profiles have the combination $(X,Y,Z) = (G,G,A)$ whereas in the AFR table `r unname(AFR_tab["GGA"])` profiles have that combination. Thus, the two populations seems to differ quite substantially.

Our goal is to test a profile in AFR for being an outlier in EUR. The first step is to model the interaction graph for the variables $X, Y$ and $Z$ with a decomposable graph. We use the `efs` function to fit the graph and use the `igraph` package to plot it.

```{r fig.align = "center", fig.cap = "\\label{fig:DG_EUR} $G$: Interaction graph for EUR.", fig.width = 7}
library(igraph)
DG_EUR <- efs(df_EUR, trace = FALSE)
G_adj  <- adj_list(DG_EUR)   # Used for input to the outlier model
G_A    <- adj_matrix(DG_EUR) # Used for igraph plots
iG_A   <- igraph::graph_from_adjacency_matrix(G_A, mode = "undirected")

## The arguments to the plot function are included
## to get a feel of how igraph can be used to plot graphs
set.seed(7)
plot(iG_A,
     edge.arrow.size=1, 
     vertex.label.cex=1.5, 
     vertex.label.family="Helvetica",
     vertex.label.font=1,
     vertex.shape="circle", 
     vertex.size=40, 
     vertex.label.color="blue", 
     edge.width=5
)
```

Notice that $G$ is in fact a tree (no cycles). The cliques and separator of $G$ are given by
$$
C_{1} = \{ X, Z\}, \quad C_{2} = \{Y, Z \} \quad \mbox{and} \quad S = \{ Z \}.
$$

and we can now form the _marginal tables_. The marginal table of clique $C_1$ is

```{r XZ_tab, include = FALSE}
XZ_tab <- sptable(as.matrix(df_EUR[, c("X", "Z")]))
```

```{r}
sptable(as.matrix(df_EUR[, c("X", "Z")]))
```
and for $C_2$

```{r YZ_tab, include = FALSE}
YZ_tab <- sptable(as.matrix(df_EUR[, c("Y", "Z")]))
```

```{r}
sptable(as.matrix(df_EUR[, c("Y", "Z")]))
```
and for $S$

```{r Z_tab, include = FALSE}
Z_tab <- sptable(as.matrix(df_EUR[, "Z"]))
```

```{r}
sptable(as.matrix(df_EUR[, "Z"]))
```


We can then count the number of profiles with a specific combination for the cliques and the separator. The number of profiles having the combination $(X, Z) = (G,A)$ for clique $C_1$ is for example $n_{XZ}(G,A) = `r unname(XZ_tab["GA"])`$. The number of profiles having combination $(Y, Z) = (G,A)$ for clique $C_2$ is \({n_{YZ}(G,A) = `r unname(YZ_tab["GA"])`}\) and for the separator the number of profiles with $Z = A$ is $n_Z(A) = `r unname(Z_tab["A"])`$. It can be shown that the _joint probability_ of having the combination $(X, Y, Z) = (G, G, A)$ is given by

```{r computation,  include = FALSE}
comp <- ( unname(XZ_tab["GA"]) * unname(YZ_tab["GA"]) ) / ( nrow(df_EUR) * unname(Z_tab["A"]) ) 
```

\begin{equation*}
  \frac{n_{XZ}(G,A) \cdot n_{YZ}(G,A) }{ 1006 \cdot n_{Z}(A) } = \frac{ `r unname(XZ_tab["GA"])` \cdot `r unname(YZ_tab["GA"])` }{ 1006 \cdot `r unname(Z_tab["A"])`} = `r comp`,
\end{equation*}

where `r nrow(df_EUR)` is the total number of profiles in `df_EUR`.

In other words, with low probability Europeans have the specific combination $(X,Y,Z) = (G, G, A)$ for this specific microhap under the model. If a new observed profile, say $z_{new}$, have this combination for the selected microhap it supports the belief that this profile might not originate from Europe. We can carry out the same calculations for all microhaps and combine the result into a _likelihood_ $L_0$ that expresses how likely it is that $z_{new}$ originates from Europe. We can also specify an alternative likelihood, $L_{1}$, specifying how likely it is that $z_{new}$ do not originate from Europe. We then define the _likelihood ratio_ as

\begin{equation*}
  \label{eq:LR}
  LR = \frac{L_{0}}{L_{1}},
\end{equation*}
which can be shown to be completely specified through the counts of profiles in cliques and separators for the given interaction graph. We can therefore test if $z_{new}$ is an outlier in Europe by calculating $LR$ and determine if the value of $LR$ (soon we modify $LR$ and call it _deviance_) is "too large" in which case we would reject that $z_{new}$ originates from Europe.

Under the null hypothesis, $z_{new}$ belongs to the hypothesised data base (EUR in this case) and so we append the profile to `df_EUR`. Assume $z_{new}$ is the $7'$th profile from AFR.

```{r}
z_new <- unlist(df_AFR[7, ]) # It is important to unlist here!
df_EUR_z_new <- df_EUR %>%
  bind_rows(z_new)
```

Lets see what the specific value of $z_{new}$ is 

```{r}
unname(z_new)
```

So we already have an idea, that $z_{new}$ could be an outlier in `df_EUR`. All ingredients are now given in order to make the outlier test.

```{r outliermodel, cache = TRUE}
set.seed(7)
om <- as.matrix(df_EUR_z_new) %>%
  outlier_model(G_adj, nsim = 10000)
```

What we have done is, that we have simulated $10000$ profiles from `df_EUR_7` and for each of those simulations we can calculate $LR$. In effect, we obtain an estimated _distribution_ of $LR$. More precise we get the distribution of $-2\cdot \log LR$ which is called the _deviance_. We can plot the distribution of the deviance as 

```{r fig.width = 7}
pmf(om)
```

Since we only have \(3\) variables it looks a bit odd. However, the particular deviance score for \(z_{new}\) is given by

```{r}
deviance(om, z_new)
```

with p-value for $z_{new}$ being an outlier in EUR

```{r}
p_val(om, deviance(om, z_new))
```

Therefore we declare $z_{new}$ as an outlier in EUR on a significance level $\alpha = 0.05$.

### A Larger Example
In this example we include more variables from `tgp_dat` and show how to aggregate decomposable graphs using expert knowledge. We know, that the SNP variables are _independent_ between microhaplotypes. We include the first $25$ microhaplotypes which consists of $96$ SNP variables.

```{r}
microhaps <- tgp_haps[1:25]
vars   <- unname(unlist(microhaps))

df_EUR_micro <- tgp_dat %>%
  filter(pop_meta == "EUR") %>%
  select(vars)

df_AFR_micro <- tgp_dat %>%
  filter(pop_meta == "AFR") %>%
  select(!!vars)
```

Then we fit an interaction graph on each microhaplotype

```{r}
micro_graphs <- lapply(microhaps, function(x) {
  df_x <- df_EUR_micro[, x]
  efs(df_x, trace = FALSE)
})
```

Then we extract the graphs and "aggregate" them (more precisely we take the union of the graphs)

```{r}
G_union <- micro_graphs %>%
  unname() %>%
  lapply(function(x) adj_list(x)) %>%
  unlist(recursive = FALSE)
```

Plotting the fitted interaction graph

```{r fig.width = 8, fig.height = 8, fig.align = "center"}
iG_union <- as_adj_mat(G_union) %>%
  igraph::graph_from_adjacency_matrix(mode = "undirected")

plot(iG_union,
     vertex.label.cex = 0.2, 
     vertex.size      = 2, 
     edge.width       = 1
)
```

Constructing the outlier model where we first append the hypothesized outlier from AFR

```{r large_outliermodel, cache = TRUE}
set.seed(7)
z_new <- unlist(df_AFR_micro[7, ]) # It is important to unlist here!

df_EUR_micro_z_new <- df_EUR_micro %>%
  bind_rows(z_new)

om_micro <- as.matrix(df_EUR_micro_z_new) %>%
  outlier_model(G_union, nsim = 10000)

# print model info
om_micro
```

Plotting the distribution of the deviance

```{r fig.width = 7, fig.align = "center"}
pmf(om_micro)
```

Determine if $z_{new}$ is an outlier

```{r}
dz_new <- deviance(om_micro, z_new)
dz_new
p_val(om_micro, dz_new)
```

It is seen again, that $z_{new}$ does not belong to `df_EUR` with a very high degree of certainty at a significance level of $0.05$.

<!-- ## Conditional Tables -->

<!-- ```{r} -->
<!-- sptab <- sptable(as.matrix(df_EUR_micro[, 1:5])) -->
<!-- csptable(sptab, structure(c(1, 5), names = c("G", "C"))) -->
<!-- ``` -->


# References
